### 2017-02-03 paper/chapters/introduction.tex
minor corrections (fig/eq uppercase)

\begin{itemize}
  \item \textbf{Images} consisting of millions of pixels can be interpreted as data vectors with each pixel representing an own dimension.
  However, these dimensions are high correlated and structured, for example nearby pixels have with high probability similar values.
  Looking at images of handwritten digits, the data can be reduced to a handful variables which may include the digit number, stroke width, position or size.
  Statistical models help us discover these relations and reduce them to representations in lower-dimensional space.
  %\item commonly have thousands of dimensions, but the underlying structure is often far simpler. For example are 28x28 images of handwritten digits 784-dimensional but assuming one-hot encoding the most important information - the digit - it can be explained using 10 dimensions. There are a few more dimensions, stroke width, cursive, position, size. But still far lower than 784. We will take a look at this example later on in \ref{sec:vae}.
  \item \textbf{Audio data} is omni-present in the real world, however most data sources can be transformed from raw audio samples to spoken words or musical notes without losing too much information for reconstruction.
  %\item \textbf{...}
\end{itemize}
% cope with
%To develop systems learning from real world data we use statistical models which are able to infer the underlying structure from
%Real world data is often high-dimensional and noise and transforming this data into an compact representation is therefore an important and also challenging task to perform.
%Generative models are ideal for this task since it has to understand the underlying structure to generate similar data.
%Building this structure is called \textbf{representation learning} and is key to many generative models and how they perform.\\

@@

The assumption that the observed variables $x$ can be explained using the causal factors $z$ can be expressed as graphical model shown in figure \ref{fig:dgm}.
%as having the conditional probability $p(x|z)$ where the latent variables $z$ are the causal factors. This stochastic dependency can be represented in a graphical model shown in figure \ref{fig:dgm}.\\

>>>>

\begin{itemize}
  \item \textbf{Images} consisting of millions of pixels can be interpreted as data vectors with each pixel representing an own dimension.
  However, these dimensions are highly correlated and structured, for example nearby pixels have with high probability similar values.
  Looking at images of handwritten digits, the data can be reduced to a handful of variables which may include the digit number, stroke width, position or size.
  Statistical models help us discover these relations and reduce them to representations in lower-dimensional space.
  %\item commonly have thousands of dimensions, but the underlying structure is often far simpler. For example are 28x28 images of handwritten digits 784-dimensional but assuming one-hot encoding the most important information - the digit - it can be explained using 10 dimensions. There are a few more dimensions, stroke width, cursive, position, size. But still far lower than 784. We will take a look at this example later on in \ref{sec:vae}.
  \item \textbf{Audio data} is omni-present in the real world, however most data sources can be transformed from raw audio samples to spoken words or musical notes without losing too much information for reconstruction.
  %\item \textbf{...}
\end{itemize}
% cope with
%To develop systems learning from real world data we use statistical models which are able to infer the underlying structure from
%Real world data is often high-dimensional and noise and transforming this data into an compact representation is therefore an important and also challenging task to perform.
%Generative models are ideal for this task since it has to understand the underlying structure to generate similar data.
%Building this structure is called \textbf{representation learning} and is key to many generative models and how they perform.\\

@@

The assumption that the observed variables $x$ can be explained using the causal factors $z$ can be expressed as graphical model shown in Figure \ref{fig:dgm}.
%as having the conditional probability $p(x|z)$ where the latent variables $z$ are the causal factors. This stochastic dependency can be represented in a graphical model shown in figure \ref{fig:dgm}.\\

### 2017-02-03 paper/chapters/sigmoid_belief_nets.tex
minor corrections (fig/eq uppercase)

Sigmoid Belief Networks (SBN) were proposed in 1992 by R.M. Neal \cite{neal:1992} representing a specific type of bayesian networks \cite{pearl:1985} shown in figure \ref{fig:sbn_arch}
In the case of SBNs however, all states are binary and the activation function is the sigmoid function. They are one of the earliest neural networks used for generative modeling, predating all other presented models in this article.
%As described in the overview above, SBNs contain a number of binary states $s$, most commonly divided into many layers.
Each state $s_i$ is a random variable in the bayesian framework which means it can be observed, hidden or unknown representing different states of knowledge. As such it can also be interpreted with a probability distribution.
\begin{equation}
  \label{eq:sbn_node}
p(s_i) = \sigma\bigg(\sum_{j<i}W_{i,j}s_j+b_i\bigg)
\end{equation}
In equation \ref{eq:sbn_node}, $\sigma$ denotes the sigmoid function $\frac{1}{1 + e^{-x}}$ and $W_{i,j}$ the connection weights from state $s_j$ to $s_i$.

To draw samples from this network, it is possible to simply evaluate equation \ref{eq:sbn_node} for each visible node.
However prior to that, the SBN first needs to learn how to generate good samples which includes the \emph{learning} and the \emph{inference} problem.\\
Learning means in this context to adjust the weights and biases of the network to be more likely to generate previously observed input.
Meanwhile statistical inference is tasked to infer the states of the hidden nodes based on the states in the visible layer.\\

>>>>

Sigmoid Belief Networks (SBN) were proposed in 1992 by R.M. Neal \cite{neal:1992} representing a specific type of bayesian networks \cite{pearl:1985} shown in figure \ref{fig:sbn_arch}
In the case of SBNs however, all states are binary and the activation function is the sigmoid function. They are one of the earliest neural networks used for generative modeling, predating all other presented models in this article.
%As described in the overview above, SBNs contain a number of binary states $s$, most commonly divided into many layers.
Each state $s_i$ is a random variable in the bayesian framework which means it can be observed, hidden or unknown representing different states of knowledge. As such it can also be interpreted with a probability distribution.
\begin{equation}
  \label{eq:sbn_node}
p(s_i) = \sigma\bigg(\sum_{j<i}W_{i,j}s_j+b_i\bigg)
\end{equation}
In Equation \ref{eq:sbn_node}, $\sigma$ denotes the sigmoid function $\frac{1}{1 + e^{-x}}$ and $W_{i,j}$ the connection weights from state $s_j$ to $s_i$.

To draw samples from this network, it is possible to simply evaluate Equation \ref{eq:sbn_node} for each visible node.
However prior to that, the SBN first needs to learn how to generate good samples which includes the \emph{learning} and the \emph{inference} problem.\\
Learning means in this context to adjust the weights and biases of the network to be more likely to generate previously observed input.
Meanwhile statistical inference is tasked to infer the states of the hidden nodes based on the states in the visible layer.\\

### 2017-02-03 paper/chapters/vae.tex
minor corrections (fig/eq uppercase)

The probabilistic encoder generally implemented as feed-forward network receives input from the dataset and outputs parameters of a distribution. For example the parameters for a gaussian distribution with diagonal covariance, in which case the networks emits the mean $\mu$ and $\sigma$ used to define the gaussian distribution $\mathcal{N}(\mu,\sigma \times I)$, where $I$ is the identity matrix.\\

@@

Properly defined emits the described probabilitic encoder variational parameters $\phi$ used to define the probability distribution $q_\phi(z|x)$.
Equally for the probabilistic decoder which constructs a distribution $p_\theta(x|z)$ with the parameters $\theta$.
This architecture is shown in figure \ref{fig:vae_architecture} with a one-dimensonal gaussian prior.
\\

@@

These intractable posteriors unfortunately arise in almost all interesting models and there are mostly two different approaches to approximate the posterior distribution.
% MCMC
Using Markov chain Monte Carlo (MCMC) methods, which construct a Markov chain with the desired attributes matching those of the posterior. The advantage of this approach is that is guaranteed to asympotically approach the desired posterior, however it is difficult to reason about the convergence rate. For deep neural networks, running MCMC in the inner loop of optimization techniques is impractical.\\
% variational inference
Using variational inference is the second approach, used in the variational autoencoder, which picks a specific distribution restricted to a family of distributions matching the posterior most closely.
Advantages using this approach rather than MCMC is the greater speed at which the computation is performed.
However it is not guaranteed to approximate the posterior accurately, if the true posterior can not be approximately expressed as a distribution of the family.

@@

\subsubsection{Kullback-Leibler Divergence}
In order to approximate one distribution $Q$ to another distribution $P$, there needs to be a measurement how close $Q$ is to $P$.
One commonly used measurement is called Kullback-Leibler divergence (short KL divergence) which enables the variational inference methods to choose appropiate distributions.
%\paragraph{The Kullback-Leibler Divergence} (KL divergence) is a measurement for the difference between two probability distributions. $\mathcal{D}_{\mathrm{KL}}(P \| Q)$ can informally described as the amount of information which is lost when using $Q$ to represent $P$.
For discrete probability distributions $P$ and $Q$, the KL divergence from $Q$ to $P$ is defined\cite{kl_div:1951} in equation \ref{eq:kl_div_disc}:
\begin{equation}
  \label{eq:kl_div_disc}
  \mathcal{D}_{\mathrm{KL}}(P || Q) = \sum_{i} P(i) \log \frac{P(i)}{Q(i)}
\end{equation}
For continuous distributions $P$ and $Q$ the KL divergence is defined similarly in equation \ref{eq:kl_div_cont}.
\begin{equation}
  \label{eq:kl_div_cont}
  \mathcal{D}_{\mathrm{KL}}(P || Q) = \int P(i) \log \frac{P(i)}{Q(i)} di
\end{equation}

@@

However we are not able to optimize equation \ref{eq:vae_obj} directly, as the intractable evidence $p(x)$ appears during the expansion of the posterior $p(z|x)$ to $\frac{p(x|z)p(z)}{p(x)}$.\\
The denominator $p(x)$ can be computed by marginalizing out the hidden variables $z$: 
\begin{equation}
  \label{eq:integral}
p(x) = \int p(x,z) p(z) dz
\end{equation}

Computing the integral in equation \ref{eq:integral} requires the evaluation of every possible configuration over the hidden variables $z$ and is therefore intractable.\\\\

@@

% more math:
% http://stats.stackexchange.com/questions/199605/how-does-the-reparameterization-trick-for-vaes-work-and-why-is-it-important
To be able to backpropagate the loss function through the VAE, it has to be differentiable and deterministic.
However, because we only sample from the constructed distribution $q_\phi(z|x)$, we cannot backpropagate through the node due to the stochastic nature.
In order to circumvent this restriction, the so-called "reparameterization trick" is applied.
Instead of drawing $z$ directly from $q_\phi(z|x)$, we first sample first an auxiliary variable $\epsilon$ from the isotropic gaussian $\mathcal{N}(0, I)$. Then we apply equation \ref{eq:rep_trick} to transform $\epsilon$ into $z$.
\begin{equation}
  \label{eq:rep_trick}
  z = \mu(x) + \Sigma^{1/2}(x)*\epsilon
\end{equation}
Note that this transformation is entirely deterministic and as such allows for backpropagation through the transformation into the variational parameters $\phi$.

@@

\subsection{Experiments}
In order to evaluate the inference performance and explore the learned latent space, the VAE has been implemented and trained\footnote{The source code is available on GitHub: \url{https://github.com/Spotlight0xff/deeplearning-seminar}}. Results of this experiments are shown in fig. \ref{fig:vae_experiments}.
%The model has been trained on the MNIST dataset \cite{mnist} with satisfactory generative performance as shown in figure \ref{fig:vae_experiments}

>>>>

The probabilistic encoder generally implemented as feed-forward network receives input from the dataset and outputs parameters of a distribution. For example the parameters for a gaussian distribution with diagonal covariance, in which case the network emits the mean $\mu$ and $\sigma$ used to define the gaussian distribution $\mathcal{N}(\mu,\sigma \times I)$, where $I$ is the identity matrix.\\

@@

Properly defined emits the described probabilitic encoder variational parameters $\phi$ used to define the probability distribution $q_\phi(z|x)$.
Equally for the probabilistic decoder which constructs a distribution $p_\theta(x|z)$ with the parameters $\theta$.
This architecture is shown in Figure \ref{fig:vae_architecture} with a one-dimensonal gaussian prior.
\\

@@

These intractable posteriors unfortunately arise in almost all interesting models and there are mostly two different approaches to approximate the posterior distribution.
% MCMC
Using Markov chain Monte Carlo (MCMC) methods, which construct a Markov chain with the desired attributes matching those of the posterior. The advantage of this approach is that it is guaranteed to asympotically approach the desired posterior, however it is difficult to reason about the convergence rate. For deep neural networks, running MCMC in the inner loop of optimization techniques is impractical.\\
% variational inference
Using variational inference is the second approach, used in the variational autoencoder, which picks a specific distribution restricted to a family of distributions matching the posterior most closely.
Advantages using this approach rather than MCMC is the greater speed at which the computation is performed.
However it is not guaranteed to approximate the posterior accurately, if the true posterior can not be approximately expressed as a distribution of the family.

@@

\subsubsection{Kullback-Leibler Divergence}
In order to approximate one distribution $Q$ to another distribution $P$, there needs to be a measurement of how close $Q$ is to $P$.
One commonly used measurement is called Kullback-Leibler (KL) divergence\cite{kl_div:1951} which enables the variational inference methods to choose appropiate distributions.
%\paragraph{The Kullback-Leibler Divergence} (KL divergence) is a measurement for the difference between two probability distributions. $\mathcal{D}_{\mathrm{KL}}(P \| Q)$ can informally described as the amount of information which is lost when using $Q$ to represent $P$.
The KL divergence from $Q$ to $P$ is defined in Equation \ref{eq:kl_div}.
\begin{equation}
  \label{eq:kl_div_disc}
  \mathcal{D}_{\mathrm{KL}}(P || Q) = \sum_{i} P(i) \log \frac{P(i)}{Q(i)}
\end{equation}
For continuous distributions $P$ and $Q$ the KL divergence is defined similarly in equation \ref{eq:kl_div_cont}.
\begin{equation}
  \label{eq:kl_div_cont}
  \mathcal{D}_{\mathrm{KL}}(P || Q) = \int P(i) \log \frac{P(i)}{Q(i)} di
\end{equation}

@@

However we are not able to optimize Equation \ref{eq:vae_obj} directly, as the intractable evidence $p(x)$ appears during the expansion of the posterior $p(z|x)$ to $\frac{p(x|z)p(z)}{p(x)}$.\\
The denominator $p(x)$ can be computed by marginalizing out the hidden variables $z$: 
\begin{equation}
  \label{eq:integral}
p(x) = \int p(x,z) p(z) dz
\end{equation}

Computing the integral in Equation \ref{eq:integral} requires the evaluation of every possible configuration over the hidden variables $z$ and is therefore intractable.\\\\

@@

% more math:
% http://stats.stackexchange.com/questions/199605/how-does-the-reparameterization-trick-for-vaes-work-and-why-is-it-important
To be able to backpropagate the loss function through the VAE, it has to be differentiable and deterministic.
However, because we only sample from the constructed distribution $q_\phi(z|x)$, we cannot backpropagate through the node due to the stochastic nature.
In order to circumvent this restriction, the so-called "reparameterization trick" is applied.
Instead of drawing $z$ directly from $q_\phi(z|x)$, we first sample an auxiliary variable $\epsilon$ from the isotropic gaussian $\mathcal{N}(0, I)$. Then we apply Equation \ref{eq:rep_trick} to transform $\epsilon$ into $z$.
\begin{equation}
  \label{eq:rep_trick}
  z = \mu(x) + \Sigma^{1/2}(x)*\epsilon
\end{equation}
Note that this transformation is entirely deterministic and as such allows for backpropagation through the transformation into the variational parameters $\phi$.

@@

\subsection{Experiments}
In order to evaluate the inference performance and explore the learned latent space, the VAE has been implemented and trained\footnote{The source code is available on GitHub: \url{https://github.com/Spotlight0xff/deeplearning-seminar}}. Results of this experiments are shown in Figure \ref{fig:vae_experiments}.
%The model has been trained on the MNIST dataset \cite{mnist} with satisfactory generative performance as shown in figure \ref{fig:vae_experiments}
